---
title: "Intro to Visualziations with Legos"
author: "Jason Bryer"
date: "September 15, 2014"
output: html_document
---

### Introduction

This document introduces many of the graph types using [`ggplot2`](http://ggplot2.org/) the Lego package developed by [Sean Kross](https://github.com/seankross/lego). The `lego` package contains a data frame, `lego` containing all Lego sets from 1970 to 2014.

```{r install, eval=FALSE}
install.packages(c('devtools','ggplot2'), repos='http://cran.r-project.org')
devtools::install_github("seankross/lego")
```

```{r setup}
library(ggplot2)
library(lego)
```

```{r}
data(lego)
str(lego)
```

The `legosets` data frame contains `r nrow(legosets)` rows with `r ncol(legosets)` columns with information about every Lego set manufactured between `r min(legosets$Year)` and `r max(legosets$Year). 

### Qualitiative Variables

#### Bar Plots

#### Side-by-Side Boxplots
=

### Quantitative Variables

#### Scatterplots

#### Dot Plots

##### Stacked Dot Plots


#### Histograms

Bin widths.

#### Boxplots



