---
title: "Lab 0: Introduction to R and RStudio"
author: "Jason Bryer"
output: html_document
---

```{r}
source("http://www.openintro.org/stat/data/present.R")
source("http://www.openintro.org/stat/data/arbuthnot.R")
```

#### 1. What years are included in this data set? What are the dimensions of the data frame and what are the variable or column names?

Column Names: 

```{r}
names(present)
```

Years:

```{r}
unique(present$year)
```

Dimensions: 

```{r}
dim(present)
```


#### 2. How do these counts compare to Arbuthnot’s? Are they on a similar scale?

```{r}
head(present)
head(arbuthnot)
```

No, these appear to be in different scales. The present dataset has births in millions, the arbuthnot has births in thousands.


#### 3. Does Arbuthnot’s observation about boys being born in greater proportion than girls hold up in the U.S.?


#### 4. Make a plot that displays the boy-to-girl ratio for every year in the data set. What do you see?

```{r}
present$boysToGirls <- present$boys / present$girls
plot(x=present$year, y=present$boysToGirls, type='l', xlab='Year', ylab='Ratio of Boys-to-Girls')
```

#### 5. In what year did we see the most total number of births in the U.S.? You can refer to the help files or the R reference card (http://cran.r-project.org/doc/contrib/Short-refcard.pdf) to find helpful commands.

```{r}
present$total <- present$boys + present$girls
barplot(present$total, names.arg=present$year)
max(present$total)
present[1,]
present[,1]
present[present$total == max(present$total),]
```

The year 1961 had the most total births with 4,268,326.

